\documentclass[12pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{geometry}
\usepackage{enumitem}
\setitemize{noitemsep}
\usepackage{tabularx}
\usepackage{setspace}
\newcolumntype{x}{>{\centering\arraybackslash}X}

\usepackage{fontspec,xunicode}
\defaultfontfeatures{Mapping=tex-text}
\setsansfont{TeX Gyre Heros}
\setromanfont{Heuristica}

\usepackage{unicode-math}
\setmathfont{TeX Gyre Schola Math}

\usepackage{titlesec} %Package for adjusting the fonts of the \section and \subsection and so forth.
\titleformat{\section}{\Large\bfseries\sffamily}{\thesection.}{1em}{}
\titleformat{\subsection}{\large\bfseries\sffamily}{\thesubsection.}{1em}{}
\titleformat{\subsubsection}{\normalsize\bfseries\sffamily}{\thesubsubsection.}{1em}{}

\onehalfspacing

\begin{document}
\global\long\def\bpi{\bar{p}}%
\global\long\def\upi{\underline{p}}%


\section{Introduction}

Humans see other humans as members of groups, and they treat them differently
based on their group membership. Cognitively, they classify humans as
members of different groups, form stereotypes about other groups, and track 
group reputations. Motivationally, they form positive or negative attitudes to 
other groups and may discriminate against some groups. They may also reciprocate 
towards groups as a whole: that is, they may respond to one person's actions by
acting differently towards his or her fellow group members.

The cognitive and motivational aspects of groups are linked. Logically, the
ability to track individuals' group membership is necessary to instantiate 
different motivations towards them.... XXX

We have experimental evidence for group reciprocity (XXX). But so far there is
no theoretical account of how it might have evolved. Group reciprocity is related
to generalized upstream reciprocity (XXX), and the literature shows that it is
hard for this to evolve except in special circumstances.

This paper sets out a theoretical model in which group reciprocity can evolve.
The requirements are quite minimal. Individuals know other individuals' group membership, and they can recall outgroup members' past behaviour towards
them. That is enough for group reciprocity to evolve in a repeated prisoner's
dilemma. So, intergroup cognition alone is enough to make intergroup
motivations evolutionarily stable.

This result comes about as follows. Groups of group-reciprocators are able to
establish cooperative relationships with other groups of group-reciprocators,
while they avoid cooperating with groups of selfish types. Selfish types within
a group of mostly group-reciprocators get the highest payoff, but this is balanced
out because most selfish types are in mostly-selfish groups, who have too
few reciprocators to sustain cooperation.

We also use computations to check the robustness of our theoretical model. (XXX
result preview.)

Group reciprocity is important in several ways. Theoretically, interethnic 
peace may be sustained by the threat of reciprocation. This could have allowed
humans to live in peace with neighbouring groups (perhaps punctuated by episodes of
punitive conflict), whereas chimpanzees always attack neighbouring bands on 
contact. It could also enable deeper forms of intergroup cooperation, such
as trade. So, the capacity for group reciprocity may be a stepping stone in the
evolution of humans as a cooperative species. 

Practically, group reciprocity may underlie modern conflicts, including 
interethnic violence (XXX interethnic riot, E G in C) and civil wars (e.g. cite Israel/Palestine paper). Fearon and Laitin 1996 XXX show how interethnic peace
may be sustained by the threat of intergroup conflict. We replicate this insight
in a different framework, using an evolutionary model rather than their repeated game model. Put another way, Fearon and Laitin show how rational forward-looking 
actors can arrive at an equilibrium where interethnic peace is sustained by the
threat of war. Our aim is to show how intergroup motivations - a form of "strong
reciprocity" (XXX cite?) - could have evolved. 

Our model shows how group reciprocity can evolve, and it also has some insights
about how its form is shaped by evolutionary pressure. The most robust forms of
group reciprocity have high thresholds of cooperation - that is, reciprocators
only cooperate with groups a high proportion of whose members previously
cooperated towards them. (These high thresholds make it less likely that selfish
free riders will invade, since only groups which are almost all reciprocators
get the benefit of mutual cooperation.) The resulting equilibrium is 
"trigger happy": it sustains a high level of cooperation, but is also sensitive 
to small amounts of defection.

\section{Literature review}

...

\section{Model}

We consider a mixed population of two types, selfish and group reciprocators (GR). At the beginning of each generation, the population randomly divides into a large number of groups of size~$G$ each.
Let~$p$ denote the population share of GR types. 
Let~$p_g$ be the proportion of GR in group~$g$, which is distributed binomially.

At every step~$t$, everybody interacts with everybody. In each pair, each individual chooses between cooperation and defection. Cooperation entails a cost~$c$ to the cooperator and a benefit~$b$ to her partner. Defection carries no costs. That is, each pair plays the following Prisoner's Dilemma game:
\begin{center}
    \begin{tabularx}{0.5\textwidth}{|x|x|x|}
        \hline
        &   Cooperate   &   Defect  \\
        \hline
        Cooperate   &   $b-c$   &   $-c$    \\
        \hline
        Defect  &   $b$ &   $0$   \\
        \hline
    \end{tabularx}
\end{center}

Selfish types always defect, A GR individual~$i$ starts by cooperating, and then cooperates with all individuals belonging to group~$g$ with a probability~$\phi(l_{gi})$, where~$l_{gi}$ is the proportion of individuals from group~$g$ who cooperated with individual~$i$ in round $t-1$.
$\phi(\cdot)$ is monotonically weakly increasing. We consider the cutoff strategy: 
$$
    \phi(l_{gi}) =
    \begin{cases}
        1   &   \text{if } l_{gi} \geq k  \\
        0   &   \text{otherwise.}
    \end{cases}
$$
The fitness is the payoff at the limit where~$t\to\infty$. Equivalently, since the game always settles to a stationary action profile, it is the average payoff of~$T$ rounds when~$T\to\infty$.

Given this behaviour, GR in all groups where $p_g \geq k$ help all individuals in other groups where $p_g \geq k$ and defect against members of all other groups. GR in other groups always defect.
%
Individuals' fitness therefore depends only on whether they are in a ``supraliminal" group with $p_g \geq k$, and on their type. Let~$q$ be the 
proportion of supraliminal groups. Let~$\bpi$ be the proportion of 
GR individuals in supraliminal groups (out of the total population in such groups). Let~$\upi$ be the proportion of GR individuals in subliminal groups (out of the total population in such groups). It follows that
\begin{itemize}
    \item Group reciprocators in supraliminal groups get a payoff of~$\bpi qb - qc$.
    \item Selfish types in supraliminal groups get~$\bpi qb$.
    \item Group reciprocators and selfish types in subliminal groups get~$0$.
\end{itemize}

After each generation, reproductive success is proportional to fitness, the total population size stays the same, and children are remixed randomly into new groups of the same size.
%
The mean fitness of the GR type is

\begin{equation*}
\frac{
  \bpi q(q(\bpi b - c))
}{
  p
}
\end{equation*}

and the mean fitness of selfish types is

\begin{equation*}
\frac{
  (1 - \bpi)q(q\bpi b)
}{
  1 - p
}
\end{equation*}

After rearranging, the mean fitness of reciprocators is higher if


\begin{equation}
\label{gr-wins}
\frac{ \bpi - p}{ 1 - p} \ge \frac{c}{b}.
\end{equation}

where 

\[
\bpi = E[p_g | p_g \ge k] = \frac{1}{G} \frac{\sum_{l=kG}^G l \mathrm{Binom}(l,G,p)}{\sum_{l=kG}^G \mathrm{Binom}(l,G,p)}
\]

The LHS of \eqref{gr-wins} is decreasing in $p$ and is equal
to the threshold $k$ when $p = 0$.\footnote{
    When the share of GR in the population is very small, the probability that$p_g = k$ conditional on~$p_g\geq k$ goes to one.
}
%
It follows that there is a unique ESS with a positive share of group
reciprocators if and only if $k > \frac{c}{b}$. Otherwise the population is homogeneously selfish in the unique ESS.

\end{document}
%%% THE REST OF THE ATTEMPTS TO PROVE ARE IN THE SEPARATE FILE

\subsection*{Proof that \eqref{gr-wins} goes to $k$ as $p \to 0$}


Consider

\[
\frac{\bpi - p}{1-p}
\]
The limit as $p \to 0$ is given by $\bpi$. For $k=0$ this is
0 since $\bpi = p$. For $k>0$, write
\[
\bpi=\frac{1}{G}\frac{\sum_{l=kG}^{G}l\phi(l,G,p)}{\sum_{l=kG}^{G}\phi(l,G,p)}=\frac{1}{G}\frac{\sum_{l=kG}^{G}l\binom{G}{l}p^{l}(1-p)^{G-l}}{\sum_{l=kG}^{G}\binom{G}{l}p^{l}(1-p)^{G-l}}.
\]

The limit of this is given by L'H\^{o}pital's rule, since the denominator
goes to zero. Differentiating top and bottom gives
\[
\frac{1}{G}\frac{\sum_{l=kG}^{G}l\binom{G}{l}(l-Gp)p^{l-1}(1-p)^{G-l-1}}{\sum_{l=kG}^{G}\binom{G}{l}(l-Gp)p^{l-1}(1-p)^{G-l-1}}
\]
so we have to apply the rule again. Continuing thus we eventually hit:
\[
\frac{1}{G}\frac{kG\binom{G}{l}(l-Gp)(l-1-Gp)(\cdots)(l-kG-Gp)p^{0}(1-p)^{G-2kG}+\cdots}{\binom{G}{l}(l-Gp)(l-1-Gp)(\cdots)(l-kG-Gp)p^{0}(1-p)^{G-2kG}+\cdots}
\]
 where all but the first terms vanish, leaving
\[
\frac{1}{G}\frac{kG\binom{G}{l}(l-Gp)(l-1-Gp)(\cdots)(l-kG-Gp)p^{0}(1-p)^{G-2kG}}{\binom{G}{l}(l-Gp)(l-1-Gp)(\cdots)(l-kG-Gp)p^{0}(1-p)^{G-2kG}}=\frac{1}{G}kG=k.
\]


(Old stuff)

\begin{align}
    q & = & 1 - CDF(G, p') \\
    \bpi & = & E[p_k | p_k \ge c] \\
    \upi & = & E[p_k | p_k < c]
\end{align}

Where $CDF$ is the binomial distribution of $p'$ out of $G$ successes, and
the conditional expectations are also taken according to this distribution.

Computations are in continuous-model-computations.R in dropbox. Plotting
reveals:

When $p$ is sufficiently below $c$, $p' = p$ and group reciprocity is 
selectively neutral, since group reciprocators and selfish types are
almost all in subliminal groups.

For some parameters, and intermediate values of $p$, $p'>p$ so that
group reciprocity is being selected for. That is, some groups successfully
manage to cooperate with other groups and have higher average fitness.

For high enough values of $p$, $p' < p$. This is because when almost everyone
is cooperating, a selfish type is almost surely in a supraliminal group
and benefits by freeriding.

Conjectures and questions.

1. What if the threshold can evolve? Might higher thresholds always 
be beneficial? If so, probably only few group reciprocators exist,
and very few are in supraliminal groups. One way to do this is
generalize the model to have high- and low-threshold types, and see
who wins. Note that in the model as is, a selfish type just has a
"threshold" of above 1. Another approach is to ask if a single
person with a different threshold can invade. He will change the
behaviour of other groups to him, but as there are many groups,
we can ignore the effects on other people's fitness and just
ask if he outcompetes other group reciprocators.
\end{document}
